Methods for Implementing Integrated Step-Selection Functions with Incomplete Data

### 2024-07-01 Chapter_4/04_Manuscript/Manuscript.tex
Updated manuscript
Minor updates
Publish PHD

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
% \usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage[colorinlistoftodos]{todonotes}

%------------------------------------------------------------------------------
%	Some Styling
%------------------------------------------------------------------------------
% Creating some TikZ styles
\tikzset{
  nonterminal/.style = {rectangle
    , minimum size = 6mm
    , very thick
    , draw = black!
  }
}

@@

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Applying Step Selection Functions to Data with Data Gaps: Simulation and
Case Study on African wild dogs}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \and
  Gabriele Cozzi\textsuperscript{1,2} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \and
  Dominik M. Behr\textsuperscript{1,2}
}

@@

\textsuperscript{1} Department of Evolutionary Biology and Environmental
Studies, University of Zurich, Winterthurerstarsse 190, 8057 Zurich,
Switzerland.

\textsuperscript{2} Botswana Predator Conservation Program, Private Bag 13,
Maun, Botswana.

\textsuperscript{\S} Corresponding author (david.hofmann2@uzh.ch)

@@

\textbf{Running Title:} None

@@

\textbf{Keywords:} step selection functions, gps data, animal movement, Lycaon
pictus, irregular data

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Animal movement data collected in the field is often noisy and incomplete, thus
confronting movement ecologists with temporal irregularities in their data. Step
selection functions have become a popular tool to analyse movement data and to
infer a species movement or preferences, yet they are based on the assumption of
equal temporal gaps between consecutive datapoints. To adhere to this
requirement, researchers typically only consider bursts of consecutive
relocations with equal step-durations, thus discarding large portions of their
data. Here, we use simulated data to question this practice and we examine
whether including temporally irregular can improve estimates from step-selection
models. In addition, we explore the usefulness of multiple methods to account
for temporal irregularity in movement data. Our results show that including
irregular data improves model estimates substantially. By applying methods that
account for the irregularity in the data, the precision of model estimates can
further be improved, albeit only marginally.
\end{abstract}

%------------------------------------------------------------------------------
%	Main Text
%------------------------------------------------------------------------------
\newpage

\onehalfspacing
\tableofcontents
\doublespacing

@@

% Create linenumbers
% \linenumbers

\section{Introduction}
GPS data is usually collected on fixed temporal intervals (could do a quick and
dirty analysis on MoveBank) ranging from a few minutes, to multiple hours or
even days between subsequent reolocations. Nevertheless, GPS devices regularly
fail to adhere to the anticipated fixrate schedule due to cloud cover or other
unfavorable conditions, thus resulting in irregular data. In other cases, the
GPS schedule is not strictly fixed but may be subject to change. For example,
\cite{Cozzi.2020} increased the fixrate of their GPS collars once per week from
4 to 8 fixes per day, aiming to collect more fine scaled data for a limited
time. We chose to coin this type of data \textit{irregular}, because it is not
collected on the anticipated fixrate schedule and therefore not directly
comparable to the vast majority of the collected data.

Step selection analysis is a powerful framework that enables researchers to
study habitat and movement preferences of their focal species. The method works
by comparing spatial covariates at locations where the studied animal has been
observed, to a set of locations where the animal was presumably absent. For
this, the collected GPS data is converted into steps, where a step represents
the straight line movement between two consecutive GPS fixes. One can then
compare covariates experienced along observed steps to covariates along
potential alternative steps and thus infer preferred or avoided features. By
restricting the availability domain to a set of steps, SSA is specifically
intended to account for spatio-temporal autocorrelation inherent to data
collected using GPS collars and is therefore one of the preferred methods for
analysing such data. However, it is considered good practice to only consider
steps with similar step durations in step-selection models. Depending on the
amount of available data and depending on the number of missing GPS fixes, this
practice might entail little sacrifice. In some situations, however, removing
irregular data implies a substantial loss of information. Additionally, recent
developments in the SSA realm have brought forward several improvements to SSA
that might allow to relax the assumption of regular step-durations. This
includes the \textit{integrated} SSA approach presented by \citep{Avgar.2016},
in which inference on habitat \textit{and} movement preferences are possible
thanks to the inclusion of movement descriptors in the respective model. More
recently, the method has been further refined by \citep{Munden.2021}, who coined
the term \textit{time-varying integrated} SSA. This method was developed for
high frequency data which has been rarified using a change-point detection
algorithm, thus also resulting in irregular step-durations. In case of missing
steps, in contrast, steps durations are not entirely random but usually
clustered.

Here, we questioned the practice of removing any irregular data in SSA and
investigated whether such data could be used to inform step selection models.
For this, we conducted a simulation study where we simulated movement
trajectories with known preferences across a virtual landscape. We then
artifically rarified the ``observed'' data by removing GPS fixes and we analysed
this data using SSA. This allowed us to investigate if and how the inclusion of
irregular data affected model estimates. We also varied the degree of
irregularity in the data and we tested for the effects of adjusting the
parametric step-length and turning distributions to different step-durations.
Besides the simulation study, we also analysed a dataset collected on dispersing
African wild dogs and examined the implications considering or discarding
irregular data.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_NumberOfSteps.png}
  \caption{}
  \label{Covariates}
  \end{center}
\end{figure}

We hypothesized that the precision of model estimates would decrease as we
increased the missigngess in the data. We also expected that the inclusion of
irregular data would not improve the precision of estimates and rather lead to
biased values. However, we anticipated that such biases, at least in estimated
movement kernel parameters, could be reduced by appropriately adjusting the
availability domain.

\section{Methods}
\subsection{Simulation Study}
\subsubsection{Spatial Covariates}
We simulated a virtual landscape comprising three spatial covariate layers, each
with a resolution of 300 x 300 pixels (\Cref{Covariates}) spanning across x- and
y-coordinates from 0 to 300. The first layer (\textsf{water}) represented
water-bodies and was simulated using a random cluster nearest‐neighbour neutral
landscape model \citep{Saura.2000}, with the cluster-proportion set to 0.5 and
the patch occupancy of water fixed to 20\%. The second layer (\textsf{elev})
resembled an elevation layer and was simulated using a Gaussian random fields
neutral landscape model \citep{Schlather.2015}, with an autocorrelation range of
10, magnitude of variation of the landscape of 1, and a magnitude of variation
in scale of 0. We simulated both of these layers using the r-package {\tt NLMR}
\citep{Sciaini.2018}. The third layer (\textsf{dist}) simply indicated the
distance (in pixels) to the center of the virtual landscape \((x = 150, y =
150)\), and can be understood as the distance to the center of an animal's
home-range. We computed spatial distances using the r-package {\tt raster}
\citep{Hijmans.2022}. We normalized the values from all layers to a range
between zero and one.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Covariates.png}
  \caption{Virtual landscape across which we simulated movement trajectories.
  All layers have a resolution of 300 x 300 pixels and were generated randomly.
  Simulated individuals were initiated within the white dashed rectangle, which
  ensured that they would not be released directly at a map border.}
  \label{Covariates}
  \end{center}
\end{figure}

\subsubsection{Movement Simulation}
To simulate movement across the above generated virtual lanscape, we employed an
``inverted'' step-selection function, which proceeded as follows. First, we
generated a random starting location. To prevent starting points in the vicinity
of map borders, we restricted sampled locations to x- and y-coordinates between
50 and 250 (white dotted rectangle in \Cref{Covariates}). Second, we generated a
set of 10 random steps originating from the sampled starting point. We then
generated random steps by sampling turning angles from a von Mises distribution
with concentration parameter \(\kappa = 0.5\) and location parameter \( \mu = 0
\), and step lengths from a gamma distribution with shape parameter \(k = 3 \)
and scale parameter \(\theta = 1\). Third, we extracted covariate values along
each random step from the underlying covariate layers and computed the average
of each covariate along every step. Fourth, we predicted for each step the
probability of being chosen by applying \Cref{EQ1}. The vector of relative
habitat preferences used to make predictions was set to \(\beta_{water} = -1\),
\(\beta_{elev} = 0.5\), and \(\beta_{dist} = -15\). Fifth, we sampled one of the
random steps based on predicted probabilities and computed the new position of
the simulated individual. We the repated these steps until a total of 100 steps
were realized and we replicated the simulation 100 times, thus resulting in 100
independent movement trajectories (\Cref{Simulations}). Note that by excluding
interactions among habitat covariates and step metrics we implicitly assume
independent habitat and movement kernels \citep{Avgar.2016}. Although
measurement error would have been straight forward to implement, this was not
the main focus of our analysis and so we assumed GPS locations to be 100 percent
accurate.

\begin{figure}
  \begin{center}
  \includegraphics[width = 0.5\textwidth]{99_Simulations.png}
  \caption{100 Simulated movement trajectories, each comprising of 100 steps.
  Simulated individuals were initiated within the blue rectangle to mitigate
  edge effects.}
  \label{Simulations}
  \end{center}
\end{figure}

\subsubsection{Step Selection Analysis}
To assess the consequences of missing GPS data in when employing step-selection
functions, we analysed the simulated GPS data under different conditions
\Cref{Design}. Specifically, we varied the degree of data-missingness (0\% to
80\%), the forgiveness when analysing the data (1 to 5 fixes), and the
parametric distributions from which random steps were generated (static vs.
dynamic distributions). We replicated each combination of conditions 100 times
and computed averages and standard deviations from the resulting model
estimates.

In this context, missingness and forgiveness are complementary terms. While
missingness describes the fraction of data that has not been collected, we use
the term forgiveness to describe how much missingness a modeler is willing to
accept. A modeler with forgiveness of one, for instance, only considers steps
with a step-duration of one (i.e. fixes that are regularly spaced). In contrast,
a modeler with forgiveness of two would also accept steps with a step-duration
of two, thus pardoning a single missing fix.

The idea of the multiple imputation approach is as follows...

\paragraph{Data Rarefication}
We rarefied the simulated GPS data by randomly removing a fixed fraction of GPS
fixes. To assess the impact of different degrees of  ``missingness'', we
incrementally increased the fraction of missing fixes from 0.0 (complete
dataset) to 0.8 (80\% missing) with 10\% increments. The removal of GPS fixes
introduced irregular temporal intervals between remaining fixes; we will refer
to these intervals as ``step-times'', where the  ``step'' refers to the straight
line movement between two consecutive GPS fixes \citep{Turchin.1998}. In the
complete dataset, a step-time of one was assumed, whereas the step-time
increased by one for every fix that was missing between subsequent fixes. For
instance, a single missing fix between two other fixes directly increased the
step time from one to two.

\paragraph{Identifying Bursts and Computing Step Metrics}
We used the generated data to compute movement bursts. A movement burst
consisted of a sequence consecutive GPS fixes with step-times that did not
exceed the accepted forgiveness. For instance, if the forgiveness was one,
already a single missing GPS fix introduced a new burst. In contrast, if the
forgiveness was set to two, step-times of two units were allowed, without
introducing a new step.

We will use the term  ``forgiveness'' to indicate how many missing fixes were
accepted before a new burst was enforced. For instance, a forgiveness of one
implied that a single missing fix did not result in a new burst. We then
converted the GPS fixes belonging to the same burst into steps, where each step
resembled the straight line movement between two GPS fixes \citep{Turchin.1998}.
For each step we computed a set of step metrics. This includd the step length,
its logarithm, the turning angle and its cosine.

\paragraph{Step Selection Function}
To conduct step selection analysis, we paired each observed step with 10 random
steps. We generated random steps by sampling turning angles from a von Mises
distribution and step lengths from a gamma distribution. Depending on the study
design, we either sampled turning angles and step lengths from fixed or fixed or
dynamic distributions (cfr. Section ...). Along observed and random steps, we
extracted underlying covariates and computed their averages along steps.

\paragraph{Step Length and Turning Angle Distributions}
To generate random steps, we employed three competing approaches, coined
\textit{uncorrected}, \textit{na\"ive}, and \textit{dynamic} approach,
respectively. In the uncorrected approach, we sampled step-lengths and
turning-angles from parametric gamma and von Mises distributions that were
fitted to observed steps with step-duration of one. In the \textit{na\"ive}
approach, we sampled step lengths and turning angles from the same
distributions, yet we linearly scaled sampled step lengths to the actual
step-duration. For instance, for any step with a step-duration of two, we
doubled the sampled step length. This approach is representative of the
time-varying SSF approach proposed by \cite{Munden.2021}. Finally, in the
\textit{dynamic} approach, we sampled step lengths and turning angles from
distributions that were fit to different step-durations. That is, we fitted
separate step length and turning angles distributions to every possible step
duration. To achieve this, we subsampled the observed data to 50\% of the fixes,
which resulted in varying step-durations. We then and fitted step length and
turning angle distributions to the steps from different step-durations. We
repeated this procedure 1000 times and averaged the parameter estimates across
replicates.

\paragraph{Regression Model}
We estimated ... using conditional logistic regression, implemented using the
r-package \textsf{survival} \citep{Therneau.2021}. We followed \cite{Avgar.2016}
and \cite{Fieberg.2021} and employed \textit{integrated} SSA. That is, aside
from the habitat covariates (\textsf{water, elev, dist}), we also included
descriptors of the step length and turning angle (\textsf{sl, log(sl), cos(ta)})
in our regression model. The model call was as follows:

$$
case\_ \sim water + elev + dist + sl + log(sl) + cos(ta)
$$

\subsection{Case Study}

because wild dogs move comparably little in the hours between xx and xx, we made
the simplyfing assumption that the 8-hour period from xx to xx is comparable to
a 4-hour period.

\section{Results}
\section{Discussion}
Results from our simulation study demonstrate that the inclusion of GPS fixes
can be used to gain information on relative habitat preferences of the focal
species. While the inclusion of steps from irregular sampling schemes resulted
in substantial biases for estimates related to step metrics, we showed that
these biases can effectively be mitigated by generating step lengths and turning
angles from distributions that are separately fit to steps from different step
durations. Regardless of biases in beta coefficients for step metrics, we found
that habitat selection estimates were unbiased, even when including irregular
steps. Even more, the inclusion of irregular steps drastially reduced model
uncertainty, thus suggesting that such data can be used to gain information that
would otherwise be lost.

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{textcomp}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[flushleft]{threeparttable}
\usepackage[nomarkers, nolists]{endfloat}
\usepackage{scalerel}
\usepackage{makecell}
\usepackage{ifthen}
\usepackage{tikz}
\usepackage{lineno}
\usepackage{adjustbox}
\usetikzlibrary{svg.path}

%------------------------------------------------------------------------------
%  Bibtex vs Biblatex
%------------------------------------------------------------------------------
% While bibtex is very nice to use, it's insanely slow. I therefore work with
% bibtex until the final stage. Set the value below to "true" or "false" to pick
% which one should be used.
\newboolean{usebiblatex}
\setboolean{usebiblatex}{true}

% Depending on the question, the commands below will prepare the document
% accordingly
\ifthenelse{\boolean{usebiblatex}}{
  %-----------------------------------------------------------------------------
  %  If using biblatex
  %-----------------------------------------------------------------------------
  % Load required packages and refer to correct bibliography
  \usepackage{csquotes}
  \usepackage[
      backend=biber
    , style=apa          % To display Author-year in the text
    , natbib=true        % Necessary to use citep, etc.
    , hyperref=true      % To have clickable links in the references
    , useeditor=false    % Don't add editor to publications
    , sorting=ynt        % Sort by year, name, and title
    , uniquename=false   % Avoid that initials are put
  ]{biblatex}
  \addbibresource{Literature.bib}

  % To include author in link
  \DeclareFieldFormat{citehyperref}{%
  \DeclareFieldAlias{bibhyperref}{noformat}% Avoid nested links
  \bibhyperref{#1}}
  \DeclareFieldFormat{textcitehyperref}{%
  \DeclareFieldAlias{bibhyperref}{noformat}% Avoid nested links
  \bibhyperref{%
    #1%
    \ifbool{cbx:parens}
      {\bibcloseparen\global\boolfalse{cbx:parens}}
      {}}}
  \savebibmacro{cite}
  \savebibmacro{textcite}
  \renewbibmacro*{cite}{%
    \printtext[citehyperref]{%
      \restorebibmacro{cite}%
      \usebibmacro{cite}}}
  \renewbibmacro*{textcite}{%
    \ifboolexpr{
      ( not test {\iffieldundef{prenote}} and
        test {\ifnumequal{\value{citecount}}{1}} )
      or
      ( not test {\iffieldundef{postnote}} and
        test {\ifnumequal{\value{citecount}}{\value{citetotal}}} )
    }
      {\DeclareFieldAlias{textcitehyperref}{noformat}}
      {}%
    \printtext[textcitehyperref]{%
      \restorebibmacro{textcite}%
      \usebibmacro{textcite}}}

  % Remove editor form everything but books
  \AtEveryBibitem{%
   \ifentrytype{book}{}{%
    \clearname{editor}
   }
  }
}{
  %-----------------------------------------------------------------------------
  %  If using bibtex
  %-----------------------------------------------------------------------------
  % Load required packages and select citation style
  \usepackage[round]{natbib}
  \bibliographystyle{apalike}
}

%------------------------------------------------------------------------------
%	Some Styling
%------------------------------------------------------------------------------
% Link colors
\definecolor{linkcolor}{HTML}{000000}
\hypersetup{
  colorlinks = true,
  linkcolor  = linkcolor,
  urlcolor   = linkcolor,
  citecolor  = linkcolor,
}

@@

% Avoid skip when including text from external .tex file
\newcommand{\inputy}[1]{\input{#1}\unskip}

% For supplementary material
\newcommand{\beginappendix}{%
  \setcounter{table}{0}
  \renewcommand{\thetable}{S\arabic{table}}%
  \setcounter{figure}{0}
  \renewcommand{\thefigure}{S\arabic{figure}}%
  \setcounter{equation}{0}
  \renewcommand{\theequation}{Equation S\arabic{equation}}%
  \setcounter{section}{0}
  \renewcommand{\thesection}{A.\arabic{section}}%
}

%------------------------------------------------------------------------------
%  ORCID
%------------------------------------------------------------------------------
% Creating some TikZ styles
\tikzset{
  nonterminal/.style = {rectangle
    , minimum size = 6mm
    , very thick
    , draw = black!
  }
}

% Create the ORCID logo
\definecolor{orcidlogocol}{HTML}{A6CE39}
\tikzset{
  orcidlogo/.pic={
    \fill[orcidlogocol]
      svg{M256,128c0,70.7-57.3,128-128,128C57.3,256,0,198.7,0,128C0,57.3,57.3
        ,0,128,0C198.7,0,256,57.3,256,128z};
    \fill[white]
      svg{M86.3,186.2H70.9V79.1h15.4v48.4V186.2z}
      svg{M108.9,79.1h41.6c39.6,0,57,28.3,57,53.6c0,27.5-21.5,53.6-56.8
        ,53.6h-41.8V79.1z M124.3,172.4h24.5c34.9,0,42.9-26.5
        ,42.9-39.7c0-21.5-13.7-39.7-43.7-39.7h-23.7V172.4z}
      svg{M88.7,56.8c0,5.5-4.5,10.1-10.1,10.1c-5.6,0-10.1-4.6-10.1-10.1c0-5.6
      ,4.5-10.1,10.1-10.1C84.2,46.7,88.7,51.3,88.7,56.8z};
  }
}

% Command to create the ORCID
\newcommand\orcid[1]{\href{https://orcid.org/#1}{\mbox{\scalerel*{
\begin{tikzpicture}[yscale=-1,transform shape]
  \pic{orcidlogo};
\end{tikzpicture}
}{|}}}}

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Methods for Implementing Integrated Step-Selection Functions with
Incomplete Data}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,2,\S} \orcid{0000-0003-3477-4365} \and
  Gabriele Cozzi\textsuperscript{1,2} \orcid{0000-0002-1744-1940} \and
  John Fieberg \textsuperscript{3} \orcid{0000-0002-3180-7021}
}

@@

\textsuperscript{1} Department of Evolutionary Biology and Environmental
Studies, University of Zurich, Winterthurerstrasse 190, 8057 Zurich,
Switzerland.

\textsuperscript{2} Botswana Predator Conservation Program, Wild Entrust,
Private Bag 13, Maun, Botswana.

\textsuperscript{3} Department of Fisheries, Wildlife, and Conservation Biology,
University of Minnesota, St. Paul, MN, USA.

\textsuperscript{\S} Corresponding author: \href{mailto://david.hofmann2@uzh.ch}{david.hofmann2@uzh.ch}

@@

\textbf{Running Title:} Step-Selection Analyses with Missing Data

@@

\textbf{Keywords:} animal movement, gps data, imputation, incomplete data,
missing fixes, step-selection analyses, step-selection functions

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}

Integrated step-selection analyses (iSSAs) are versatile and powerful frameworks
for studying habitat and movement preferences of tracked animals. iSSAs utilize
integrated step-selection functions (iSSFs) to model movements in discrete time,
and thus, require animal location data that are regularly spaced in time.
However, many real-world datasets are incomplete due to tracking devices failing
to locate an individual at one or more scheduled times, leading to slight
irregularities in the duration between consecutive animal locations. To address
this issue, researchers typically only consider bursts of regular data (i.e.,
sequences of locations that are equally spaced in time), thereby reducing the
number of observations used to model movement and habitat selection. We reassess
this practice and explore four alternative approaches that account for temporal
irregularity resulting from missing data. Using a simulation study, we compare
these alternatives to a baseline approach where temporal irregularity is ignored
and demonstrate the potential improvements in model performance that can be
gained by leveraging these additional data. We also showcase these benefits
using a case study on a spotted hyena (\textit{Crocuta crocuta}).

\end{abstract}

%------------------------------------------------------------------------------
%	Main Text
%------------------------------------------------------------------------------
% \newpage
%
% \onehalfspacing
% \tableofcontents
% \doublespacing

@@

% Create linenumbers
\linenumbers

\section{Introduction}

Understanding how animals move across the landscape, what habitats they prefer,
and what resources they select are fundamental questions in movement ecology
\citep{Nathan.2008}. Thanks to recent advances in animal tracking
\citep{Cagnacci.2010, Williams.2019, Beardsworth.2022} and remote sensing
technologies \citep{Toth.2016, Rumiano.2020}, new opportunities and analytical
tools have emerged for studying how animals move and interact with their
environment \citep{Tomkiewicz.2010, Kays.2015, Nathan.2022}. Methods commonly
used to analyze animal movement data, including step-selection analyses
\citep{Fortin.2005, Thurfjell.2014, Fieberg.2021} and hidden Markov models
\citep{Michelot.2016}, require \textbf{animal locations} (terms in bold at first
occurrence are defined in \Cref{tab:Glossary}) that are collected at a constant
sampling frequency, leading to data that are equally spaced in time. Yet, it is
common to encounter missing locations in most telemetry data sets
\citep{Frair.2010, Hofman.2019, Vales.2022}, which introduces unwanted
irregularities in the duration between successive locations. Thus, there is a
need for analytical tools that enable the analysis of such data, while
mitigating potential biases arising from temporal irregularity introduced
through missing animal locations.

Step-selection analyses (SSAs) are widely used to study animals' movement
capacities and habitat-selection patterns \citep{Fortin.2005, Thurfjell.2014}.
Straight-line segments connecting consecutive animal locations, referred to as
\textbf{steps}, form the basic building blocks of the statistical likelihood in
SSAs. Specifically, SSAs model the probability $u$ of finding an individual at
location $s$ at time $t+1$, given the animal's past positions at time $t$ and
$t-1$, $s_t$ and $s_{t-1}$, respectively:

\begin{equation}
\label{EQ1}
u(s_{t+1}) = \frac{\phi(s_{t+1}, s_t, s_{t-1}; \gamma)w(x(s_{t+1}); \beta)}{\int_{s \in G}\phi(s_{t+1}, s_{t}, s_{t-1}; \gamma)w(x(s_{t+1}); \beta)ds}
\end{equation}

\noindent Here, the function $\phi$ represents an animal's \textbf{movement
kernel} which is usually expressed in terms of \textbf{step-length} and
\textbf{turning-angle} distributions, with $\gamma$ representing parameters in
these distributions. The function $w$ is the \textbf{habitat-selection function}
and reflects an animal's preferences $\beta$ for environmental characteristics
$x$ at location $s_{t+1}$. In most applications, $w$ is modeled as a log-linear
function of $x$, taking the form $w = exp(x^\top\beta)$. The integral in the
denominator of \ref{EQ1} ensures that $u$ is a proper probability distribution
(i.e., that it integrates to 1). Following \citet{Michelot.2024}, we call the
product $\phi \times w$ the step-selection function (SSF), as it highlights that
the probability of finding an animal at a certain location depends on both the
animal's movement kernel and its habitat-selection function.

Given a series of \textbf{observed steps}, finding the movement and
habitat-selection parameters that maximize the likelihood in \ref{EQ1} requires
approximating the integral in the denominator for each observed step. A variety
of numerical integration techniques can be used for this purpose
\citep{Michelot.2024}, but a common approach is to combine observed steps with
\textbf{random steps} generated by sampling step lengths and turning angles from
parametric distributions informed by the data \citep{Fortin.2005,
Thurfjell.2014}. Environmental conditions at observed steps are then contrasted
with environmental conditions at random steps in a (mixed effects) conditional
logistic regression framework \citep{Fortin.2005, Muff.2020}. To jointly
estimate parameters in $\phi$ and $w$, movement descriptors (e.g., step length
(sl), its natural logarithm (log(sl)), and the cosine of the turning angle
(cos(ta))) can be included in the conditional logistic regression model, and
their estimated coefficients can be used to update the initial (tentative)
step-length and turning-angle distributions \citep{Duchesne.2015, Avgar.2016,
Fieberg.2021}. The specific descriptors that need to be included depend on the
assumed step-length and turning-angle distributions \citep[for more details, see
Appendix C of][]{Fieberg.2021}. This approach to estimating parameters of the
SSF, termed \textit{integrated} SSA (or iSSA) by \citet{Avgar.2016}, is similar
to using importance sampling to approximate the integral in \ref{EQ1}
\citep{Michelot.2024} and is readily accessible through the \texttt{R}-package
\texttt{amt} \citep{Signer.2019}.

%In the SSF framework introduced by \citet{Fortin.2005}, the movement kernel
%$\phi$ was assumed to be known, which precluded joint estimation of $\phi$ and
%$w$ and has been shown to result in biased habitat-selection estimates
%\citep{Forester.2009}. A more robust alternative is to jointly estimate $\phi$
%and $w$ by fitting tentative distributions for step-lengths and turning-angles,
%and including descriptors of step-characteristics (i.e., step-lengths and
%turning-angles) as covariates in the statistical model \citep{Forester.2009,
%Avgar.2016}. In SSFs, the use of random steps serves to generate integration
%points that are needed to approximate the integral in \ref{EQ1} and, even
%though the true movement kernel $\phi$ is often not known a priori, it can be
%approximated using \textit{tentative} distributions \citep{Michelot.2024}. The
%use of parametric distributions serves as a means of conducting importance
%sampling and allows generating random steps that contribute better towards
%improving the approximation \citep{Michelot.2024}.

SSAs have proven extremely effective in numerous ecological studies
\citep{Thurfjell.2014}, providing insights into seasonal space use
\citep{Vales.2022, Enns.2023}, resource selection during distinct behavioral
phases \citep{Elliot.2014, Abrahms.2017, Cozzi.2018, Broekhuis.2019}, and the
effects of landscape familiarity or memory on animal movements \citep{Kim.2023}.
A model parametrized using iSSA resembles a fully mechanistic movement model
that can be used to simulate space use under novel conditions \citep{Avgar.2016,
Signer.2017, Hofmann.2023, Signer.2023}. This characteristic has made iSSAs a
useful tool for quantifying landscape resistance and identifying movement
corridors \citep{Zeller.2020, Buchholtz.2020, Hofmann.2021, Hofmann.2023}.

% Glossary of terms
\input{99_Glossary.tex}

A key assumption when conducting an iSSA is that the data have been collected at
a constant sampling frequency, thus producing \textbf{trajectories} with
\textbf{regular step durations} ($\Delta t$; \citealp{Fortin.2005,
Thurfjell.2014}). Here, we refer to such data as \textbf{regular animal
locations}, and without loss of generality, we assume the regular step duration
to be one (i.e., $\Delta t = 1$). Regular step durations ensure that
step lengths and turning angles are independent of the step duration, and
therefore, steps can be pooled when estimating movement parameters.
Since animal locations are usually obtained using automated tracking devices,
such as GPS collars programmed to record data at regular intervals, satisfying
this assumption seems straightforward. In reality, however, device limitations
often imply that some of the aspired datapoints fail to be collected, thus
introducing \textbf{missingness} and confronting researchers with
\textbf{irregular animal locations} and \textbf{irregular step durations}
\citep{Frair.2010}. In a comprehensive study, \citet{Hofman.2019} showed that
across 3,000 GPS devices and 160 species, the average success rate of obtaining
a scheduled animal location was 78\% (implying a missingess of 22\%), thus
highlighting that irregular animal locations are a frequent phenomenon in
ecological studies.

It is generally recommended that, in the case of such irregular data,
researchers should only retain \textbf{bursts} of steps with regular
step durations (possibly with some tolerance) and discard the rest
\citep{Thurfjell.2014}. We will refer to this modeling approach as having a
\textbf{forgiveness} level of one, indicating that only steps with
step durations $\Delta t = 1$ are retained for further analysis. In \texttt{R},
the \texttt{amt} package provides the function \texttt{track\_resample}
specifically for identifying bursts of steps with regular step durations
\citep{Signer.2019}. The main drawback of this approach is that it may result in
a substantial amount of data being discarded (\Cref{Overview} and
\Cref{NumberOfSteps}). For instance, consider a hypothetical trajectory in which
location 4 is missing (\Cref{Overview}a). The absence of this location prevents
the computation of a step between locations 3 and 4, as well as between
locations 4 and 5. Furthermore, without these steps, it becomes impossible to
compute a turning angle for the step between locations 5 and 6. Consequently,
the lack of a single location reduces the effective sample size, which is the
number of \textbf{valid steps}, by three. Assuming that animal locations are
missing at random, a missingness level of 25\% causes the number of valid steps
to drop by $58\%$ (\Cref{NumberOfSteps}). A modeler willing to increase their
level of forgiveness to two (i.e., allowing for inclusion of steps with $\Delta
t \leq 2$) would be able to increase the number of valid steps by $57\%$
(\Cref{Overview} and \Cref{NumberOfSteps}), therefore achieving a substantial
gain in effective sample size. The ability to capitalize on irregular data is
likely to be particularly important for applications where data are already
limited, such as, for instance, when modeling dispersing individuals
\citep{Rudnick.2012, Fattebert.2015, Cozzi.2020}. However, increasing the
forgiveness also implies that step durations of the retained steps become
irregular, thus necessitating appropriate tools to account for such
irregularity.

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_Overview.png}
  \caption{(\textbf{a}) Demonstration of how missingness affects the number of
  valid steps that can be used for step-selection analyses under different
  levels of forgiveness. The upper panel depicts a trajectory with zero missing
  locations. That is, all aspired locations were successfully collected on a
  regular interval (yielding a regular step duration of $\Delta t = 1$). This
  trajectory produces four valid steps that can be included in the iSSF model
  and one invalid step that has to be omitted because it has no turning angle
  associated with it. In the central panel, animal location 4 was not obtained,
  introducing a missingness of 16.7\%. If the modeler has a forgiveness of one,
  only a single step can be included for further analysis, as all other steps
  are invalid (either because no turning angle can be computed or because
  step durations exceed the forgiveness). If, however, the modeler exhibited a
  forgiveness of two, such as in the lower panel, a total of three steps could
  be obtained for further analysis. (\textbf{b}) Conceptual illustration of how
  increasing the forgiveness allows one to retain additional steps that can be
  used for step-selection analysis. The sequence of dots resembles the sequence
  of locations that were scheduled to be collected (e.g. using a GPS device),
  with the lines representing hypothetical steps. Because not all locations were
  successfully obtained (gray dots), there is missingness. Depending on the
  forgiveness level, already a single missing location enforces the introduction
  of a new burst, which leads to the loss of several steps. In addition, some of
  the remaining steps are invalid (dotted) because they are lacking a
  turning angle. By increasing the forgiveness, a modeler is willing to retain
  steps that exceed the regular step duration by a certain threshold, which
  enables them to obtain longer bursts and increase the number of steps that can
  be used for further analysis. In the figure, forgiveness increases from left
  to right.}
  \label{Overview}
  \end{center}
\end{figure}

\begin{figure}
  \begin{center}
  \includegraphics[width = \textwidth]{99_NumberOfSteps.png}
  \caption{Illustration of how missingness in animal location data reduces the
  number of valid steps that can be used in step-selection analyses (left panel)
  and how increasing forgiveness helps to retain additional steps that are
  otherwise omitted (right panel). At a missingness of 0, 998 valid steps can be
  obtained from the total of 1,000 animal locations. At higher missingness,
  step durations become irregular, which means that the number of valid steps
  decreases substantially. However, if the modeler is willing to increase their
  forgiveness, additional steps can be gained. The right panel shows the number
  of valid steps that is gained when increasing the forgiveness from 1 to
  2, 3, 4, and 5, respectively. Ribbons indicate the 95\%-percentile intervals
  derived from 1,000 replicates.}
  \label{NumberOfSteps}
  \end{center}
\end{figure}

Various methods have been employed in the past to address temporal
irregularities in animal location data. These may serve as valuable starting
points for developing approaches that enable the integration of irregular data
with iSSFs.

\begin{enumerate}
